---
title: "Bike Data Analysis Report"
author: "Vishal L S"
date: "2024-03-21"
output:
  pdf_document: default
  html_document: default
---


# Introduction

This report presents an analysis  to understand how casual riders and annual members use  bikes differently so that we can design a new marketing strategy to convert casual riders into annual members with compelling data insights and professional data visualizations.

# Libraries
```{r loading packages}
library(tidyverse) 
library(conflicted)
```


```{r default choices, include=FALSE}
conflict_prefer("filter", "dplyr")
conflict_prefer("lag", "dplyr")
```

# Data Collection

The analysis is based on a dataset provided by Motivate International Inc.

```{r Loading data, include=FALSE}
q1_2019 <- read_csv("Divvy_Trips_2019_Q1.csv")
q1_2020 <- read_csv("Divvy_Trips_2020_Q1.csv")
```

# Wrangle Data And Combine Into A Single File

## Compare column names each of the files

```{r col names, echo=TRUE}
colnames(q1_2019)
colnames(q1_2020)
```

## Rename columns
```{r Rename Columns, echo=TRUE}
(q1_2019 <- rename(q1_2019
,ride_id = trip_id
,rideable_type = bikeid
,started_at = start_time
,ended_at = end_time
,start_station_name = from_station_name
,start_station_id = from_station_id
,end_station_name = to_station_name
,end_station_id = to_station_id
,member_casual = usertype
))

```


## Inspect the dataframes and look for incongruencies

```{r include=FALSE}
str(q1_2019)
str(q1_2020)
```
## Convert ride_id and rideable_type to character so that they can stack correctly
```{r}
q1_2019 <- mutate(q1_2019, ride_id = as.character(ride_id)
,rideable_type = as.character(rideable_type))
```
## Stack individual quarter's data frames into one big data frame
```{r}
all_trips <- bind_rows(q1_2019, q1_2020)#, q3_2019)#, q4_2019, q1_2020)
```

## Remove lat, long, birthyear, and gender fields as this data was dropped
```{r}
all_trips <- all_trips %>%
select(-c(start_lat, start_lng, end_lat, end_lng, birthyear, gender, "tripduration"))
```

# Clean Up And Add Data To Prepare For Analysis

## Inspect the new table that has been created
```{r Inspect, echo=TRUE}
colnames(all_trips)

nrow(all_trips) 

dim(all_trips)
```


```{r Inspect 2, include=FALSE}
head(all_trips)

str(all_trips)
```


```{r summary, echo=TRUE}
summary(all_trips)
```

## There are a few problems we will need to fix!!
(1) In the "member_casual" column, there are two names for members ("member" and "Subscriber") and two names for casual riders ("Customer" and "casual"). We will need to consolidate that from four to two labels.

(2) The data can only be aggregated at the ride-level, which is too granular. We will want to add some additional columns of data -- such as day, month, year -- that provide additional opportunities to aggregate the data.

(3) We will want to add a calculated field for length of ride since the 2020Q1 data did not have the "tripduration" column. We will add "ride_length" to the entire dataframe for consistency.

(4) There are some rides where tripduration shows up as negative, including several hundred
rides where Divvy took bikes out of circulation for Quality Control reasons. We will want to delete these rides.

(5)  In the "member_casual" column, replace "Subscriber" with "member" and "Customer" with "casual"
Before 2020, Divvy used different labels for these two types of riders ... we will want to make our dataframe consistent with their current nomenclature

## Begin by seeing how many observations fall under each usertype
```{r}
table(all_trips$member_casual)
```



## Reassign to the desired values
```{r include=FALSE}
all_trips <- all_trips %>%
mutate(member_casual = recode(member_casual
,"Subscriber" = "member"
,"Customer" = "casual"))
```
## Check to make sure the proper number of observations were reassigned
```{r}
table(all_trips$member_casual)
```

## Add columns that list the date, month, day, and year of each ride,This will allow us to aggregate ride data for each month, day, or year .
```{r Date, include=FALSE}
all_trips$date <- as.Date(all_trips$started_at) #The default format is yyyy-mm-dd
all_trips$month <- format(as.Date(all_trips$date), "%m")
all_trips$day <- format(as.Date(all_trips$date), "%d")
all_trips$year <- format(as.Date(all_trips$date), "%Y")
all_trips$day_of_week <- format(as.Date(all_trips$date), "%A")
```
## Add a "ride_length" calculation to all_trips (in seconds)
```{r}
all_trips$ride_length <- difftime(all_trips$ended_at,all_trips$started_at)
```

## Inspect the structure of the columns
```{r include=FALSE}
str(all_trips)
```

## Convert "ride_length" from Factor to numeric so we can run calculations on the data
```{r}
is.factor(all_trips$ride_length)
all_trips$ride_length <- as.numeric(as.character(all_trips$ride_length))
is.numeric(all_trips$ride_length)
```
## Remove "bad" data
We will create a new version of the dataframe (v2) since data is being removed

```{r new version}
all_trips_v2 <- all_trips[!(all_trips$start_station_name == "HQ QR" | all_trips$ride_length<0),]
```
# Conduct Descriptive Analysis

## Descriptive analysis on ride_length
```{r echo=TRUE}
mean(all_trips_v2$ride_length)
median(all_trips_v2$ride_length)
max(all_trips_v2$ride_length)
min(all_trips_v2$ride_length)

summary(all_trips_v2$ride_length)
```
## Compare members and casual users

```{r echo=TRUE}
aggregate(all_trips_v2$ride_length ~ all_trips_v2$member_casual, FUN = mean)
aggregate(all_trips_v2$ride_length ~ all_trips_v2$member_casual, FUN = median)
aggregate(all_trips_v2$ride_length ~ all_trips_v2$member_casual, FUN = max)
aggregate(all_trips_v2$ride_length ~ all_trips_v2$member_casual, FUN = min)
```
## See the average ride time by each day for members vs casual users
```{r echo=TRUE}
aggregate(all_trips_v2$ride_length ~ all_trips_v2$member_casual + all_trips_v2$day_of_week,
FUN = mean)
```
## Notice that the days of the week are out of order.
```{r}
all_trips_v2$day_of_week <- ordered(all_trips_v2$day_of_week, levels=c("Sunday", "Monday",
"Tuesday", "Wednesday", "Thursday", "Friday", "Saturday"))
```
## Now, let's run the average ride time by each day for members vs casual users

```{r echo=TRUE}
aggregate(all_trips_v2$ride_length ~ all_trips_v2$member_casual + all_trips_v2$day_of_week,
FUN = mean)
```
## analyze ridership data by type and weekday
```{r echo=TRUE}
all_trips_v2 %>% 
  mutate(weekday = wday(started_at, label = TRUE)) %>%
  group_by(member_casual, weekday) %>%
  summarise(number_of_rides = n()	
            ,average_duration = mean(ride_length)) %>%
  arrange(member_casual, weekday)	
```
# Visualizations

## Let's visualize the number of rides by rider type

```{r Visualize number of riders}
all_trips_v2 %>%
  mutate(weekday = wday(started_at, label = TRUE)) %>%
  group_by(member_casual, weekday) %>%
  summarise(number_of_rides = n(),
            average_duration = mean(ride_length)) %>%
  arrange(member_casual, weekday) %>%
  ggplot(aes(x = weekday, y = number_of_rides, fill = member_casual)) +
  geom_col(position = "dodge")
```
## Let's create a visualization for average duration

```{r Visualization for average duration}
all_trips_v2 %>%
mutate(weekday = wday(started_at, label = TRUE)) %>%
group_by(member_casual, weekday) %>%
summarise(number_of_rides = n()
,average_duration = mean(ride_length)) %>%
arrange(member_casual, weekday) %>%
ggplot(aes(x = weekday, y = average_duration, fill = member_casual)) +
geom_col(position = "dodge")
```

# Conclusion


In conclusion, the Cyclistic data set project in R programming has provided valuable insights into bike-share usage patterns and user behavior we have conducted a thorough analysis of  bike_data, focusing on understanding the differences in usage patterns between annual members and casual riders. Here's a comprehensive conclusion addressing the questions posed:

1. How do annual members and casual riders use Cyclistic bikes differently?

The analysis reveals distinct usage patterns between annual members and casual riders. Annual members tend to have more consistent and frequent usage throughout the week, with slightly longer average ride durations compared to casual riders. On the other hand, casual riders exhibit more sporadic usage, with peak demand on weekends and shorter average ride durations.

2. Why would casual riders buy Cyclistic annual memberships?

Casual riders might be incentivized to purchase annual memberships if Cyclistic provides attractive benefits such as discounted rates for frequent users, exclusive access to special events or promotions, and enhanced convenience features like priority access to bikes during peak hours. Additionally, highlighting the cost-effectiveness and convenience of annual memberships compared to pay-as-you-go options could persuade casual riders to opt for memberships.

3. How can Cyclistic use digital media to influence casual riders to become members?

Cyclistic can leverage digital media platforms to effectively target and engage casual riders, encouraging them to transition into becoming annual members. Strategies may include:

* Social media campaigns: Cyclistic can run targeted advertising campaigns on platforms like Facebook, Instagram, and Twitter to promote the benefits of annual memberships, highlighting cost savings, convenience, and exclusive perks.

* Email marketing: Sending personalized email newsletters to casual riders with tailored offers and incentives for joining as annual members can be effective. Cyclistic can also use email automation to follow up with users who have previously expressed interest or participated in promotional events.

* Influencer partnerships: Collaborating with local influencers or bike enthusiasts to create sponsored content endorsing Cyclistic's annual memberships can help reach a wider audience and build credibility.

* Referral programs: Implementing referral programs where existing members can earn rewards or discounts for referring friends or family to sign up for annual memberships can incentivize word-of-mouth promotion and drive membership conversions.

By implementing targeted digital media strategies tailored to the preferences and behaviors of casual riders, Cyclistic can effectively influence them to become loyal annual members, thereby increasing revenue and fostering long-term customer relationships.







